```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr,
               tidyverse,
               cowplot,
               Biobase,
               pheatmap,
               magrittr,
               testthat,
               BioQC,
               readxl,
               ggpubr,
               stringr,
               DBI,
               foreach,
               doMC,
               data.tree,
               testit)

registerDoMC(detectCores(logical=TRUE))

devtools::load_all("/storage/home/sturm/projects/immune_deconvolution_methods")
set_cibersort_binary("../lib/CIBERSORT/CIBERSORT.R")
set_cibersort_mat("../lib/CIBERSORT/LM22.txt")

color_file = "../tables/color_scales.xlsx"
color_scales = sapply(excel_sheets(color_file), function(sheet) {
  tbl = read_excel(color_file, sheet=sheet)
  colors = tbl$color
  names(colors) = tbl$value
  colors
})

tcga_db = dbConnect(RSQLite::SQLite(), "/db/tcga.sqlite", flags = RSQLite::SQLITE_RO)
```


# Input data

In this chapter, we load and preprocess the different input datasets. 
We use a dedicated environment for each dataset in order not to trash the global namespace. 


## cell type hierarchy
We use a hierachy of immune cell types to map the cell types between different datasets. 
Here is the hierachy visualized: 

```{r hierarchy, fig.height=8, fig.width=8}
cell_type_hierarchy = new.env()
with(cell_type_hierarchy, {
  tree = immunedeconv::cell_type_tree
  SetGraphStyle(tree, rankdir = "LR")
  SetEdgeStyle(tree, arrowhead = "vee", color = "grey35", penwidth = 2)
  SetNodeStyle(tree, style = "filled,rounded", shape = "box", fillcolor = "GreenYellow", 
              fontname = "helvetica", tooltip = GetDefaultTooltip, fontcolor="black")
  plot(tree)
})


```


## Single cell data for simulated mixtures

```{r single_cell_schelker, message=FALSE, cache=TRUE}
tmp_env = new.env()
with(tmp_env, {
  load("../data/single_cell_schelker/single_cell_schelker.rda")
  # symbol_remap = read_tsv("../data/single_cell_schelker/symbol_checker.tsv") %>%
  #   as.data.frame() %>%
  #   column_to_rownames("Input")
  valid_cells = which(pData(single_cell_schelker)$cell_type!="Unknown")
  eset = single_cell_schelker[,valid_cells]
  pData(eset) %<>% rename(method_cell_type=cell_type) %>%
    inner_join(immunedeconv::cell_type_mapping %>% filter(method_dataset == "schelker_single_cell"))
  # rownames(eset) = symbol_remap[,rownames(eset)]$`Approved symbol`
  assert("no unknown cells left", !any(pData(eset)$cell_type == "Unknown"))
})
single_cell_schelker = new.env()
single_cell_schelker$eset = tmp_env$eset
single_cell_schelker$cell_types = pData(single_cell_schelker$eset)$cell_type %>% unique()


```


```{r}
cell_types = as.environment(list(
  available=pData(single_cell_schelker$eset) %>% select(cell_type) %>% distinct() %>% pull(cell_type),
  cancer=c("Melanoma cell", "Ovarian carcinoma cell"),
  other=c("Cancer associated fibroblast", "Endothelial cell")
))
cell_types$immune_cells = cell_types$available[!cell_types$available %in% c(cell_types$cancer, cell_types$other)]
```

In this study, we make use of the single cell dataset curated by [@Schelker2017]. They aggregated single cell sequencing data from different sources resulting in a set of ~12,000 single cells. They classified the cells using at set of 45 marker genes into `r length(cell_types$available)` categories:
`r length(cell_types$cancer)` cancer types (`r str_c(cell_types$cancer, collapse=", ")`),
`r length(cell_types$immune)` immune cells (`r str_c(cell_types$immune, collapse=", ")`),
`r length(cell_types$other)` other cells (`r str_c(cell_types$other, collapse=", ")`) and 
*Unknown cells* which could not have been classified unambiguously. 

Unknown cells are excluded from the downstream analysis. 

Table: the ~12,000 samples by cell type
```{r, echo=FALSE}
pData(single_cell_schelker$eset) %>% group_by(cell_type) %>% count() %>% knitr::kable()
```


```{r, echo=FALSE, fig.width=10, fig.height=10, fig.cap="tSNE-clustering of the ~12000 single cells from [@Schelker2017]. "}
pData(single_cell_schelker$eset) %>% 
  ggplot(aes(x=tsneX1, y=tsneX2, colour=cell_type)) +
           geom_point(size=1) + 
           theme(legend.position="bottom") + 
           scale_color_manual(values=color_scales$cell_types)
```

<!-- ### Attempting to distinguish between macrophages and monocytes -->
<!-- ```{r} -->
<!-- cd11b = exprs(single_cell_schelker$eset)["ITGAM",] != 0  -->
<!-- pdata = pData(single_cell_schelker$eset)  -->
<!-- pdata$ITGAM = cd11b -->
<!-- pdata %>% filter(cell_type == "Macrophage/Monocyte") %>%  -->
<!--   ggplot(aes(x=tsneX1, y=tsneX2, colour=ITGAM)) + -->
<!--    geom_point(size=1) +  -->
<!--    theme(legend.position="bottom")  -->

<!-- # pheatmap(exprs(single_cell_schelker$eset)[,pdata$cell_type == "Macrophage/Monocyte"], cluster_rows = FALSE) -->

<!-- ``` -->


## Immune cell bulk reference samples (quanTIseq training data)
```{r, message=FALSE, cache=TRUE}
immune_cell_reference = new.env()
with(immune_cell_reference, {
  tidy = read_tsv("../data/immune_cell_reference_profiles/immune_cell_reference_tidy.tsv")

  sample_description = tidy %>% 
    select(sample, method_cell_type=CellType, Study) %>% 
    distinct() %>% 
    inner_join(immunedeconv::cell_type_mapping %>% filter(method_dataset == "quantiseq_reference"))

  expr = tidy %>%
  select(sample, hgnc_id, TPM) %>%
  group_by(sample, hgnc_id) %>%
  summarise(TPM = sum(TPM)) %>%
  spread(sample, TPM)

  expr_mat = expr %>%
  as.data.frame() %>%
  column_to_rownames("hgnc_id") %>% 
  as.matrix() %>% 
  .[, sample_description$sample]
})

```



## Validation data (RNAseq + matched gold standard)
e.g. IHC, Flow cytometry, single cell

```{r}
validation_datasets = read_excel("../tables/validation-datasets.xlsx")
kable(validation_datasets)
```

### Hoeck data
8 PBMC, RNAseq+flow cytometry
```{r}
tmp_env = new.env()
with(tmp_env, {
  load("../data/hoeck/HoekPBMC_mixture.RData")
  load("../data/hoeck/HoekPBMC_gtruth.RData")
})
hoeck = new.env() 
hoeck$expr_mat = tmp_env$mix.mat
hoeck$ref = tmp_env$RefData %>% 
  as_tibble(rownames="sample") %>% 
  gather(method_cell_type, true_fraction, -sample) %>%
  inner_join(immunedeconv::cell_type_mapping %>% filter(method_dataset == "hoeck"))
```


### Schelker data 
Load the data and merge replicates.
3 ovarian cancer ascites samplese which have matched flow cytometry/single cell/bulk data. 
```{r}
schelker_ovarian = new.env()
with(schelker_ovarian, {
  bulk_replicates = read_xls("../data/single_cell_schelker/ascites_bulk_samples.xls") 

  samples = list("7873M"=c("ascites_7873_1", "ascites_7873_2"),
                 "7882M"=c("ascites_7882_1", "ascites_7882_2"),
                 "7892M"=c("ascites_7892_1", "ascites_7892_2"))
  
  bulk = lapply(samples, function(cols) {
    apply(bulk_replicates[,cols], 1, mean)
  }) %>% bind_cols() 
  
  bulk %<>% 
    mutate(gene_symbol = bulk_replicates$Row)
  
  expr_mat = bulk %>%
    as.data.frame() %>%
    column_to_rownames("gene_symbol") %>%
    as.matrix()
  
  # Load the ground truth
  
  true_cell_count = pData(single_cell_schelker$eset) %>% 
    filter(source == "ascites", donor %in% colnames(bulk)) %>% 
    group_by(donor, cell_type) %>% 
    count() %>% 
    rename(cell_count=n) %>% 
    group_by(donor) %>%
    mutate(true_fraction=cell_count/sum(cell_count)) %>% 
    arrange(cell_type, donor)

  

   ref = true_cell_count %>% select(sample=donor, cell_type, true_fraction) %>% mutate(method_dataset = "schelker_ovarian")
})

```

Single cell count for the 3 samples
```{r}

schelker_ovarian$true_cell_count %>% 
  group_by(donor) %>%
  summarise(sum(cell_count)) %>%
  kable()

schelker_ovarian$true_cell_count %>% kable()

```


### Racle data
4 metastatic melanoma, matched bulkRNAseq with FlowCytometry. 
```{r, message=FALSE}
tmp_env = new.env()
with(tmp_env, {
  ensg_hgnc = read_tsv("../data/ensemble_hgnc.txt") %>% 
  rename(gene_id = `Gene stable ID`, hgnc_symbol=`HGNC symbol`)
  racle_cyto = read_excel("../data/racle/racle2017_flow_cytometry.xlsx")
  racle_files = list.files("../data/racle/GSE93722_RAW/", full.names = TRUE)
  racle_identifiers = str_extract(racle_files, "LAU\\d+")
  racle_expression_raw = lapply(list.files("../data/racle/GSE93722_RAW/", full.names = TRUE), read_tsv) 
  gene_ids = racle_expression_raw[[1]]$gene_id
  racle_tpm = sapply(racle_expression_raw, function(df) {pull(df, "TPM")}) %>% as.tibble()
  colnames(racle_tpm) = racle_identifiers
  racle_tpm$gene_id = gene_ids
  racle_expression = racle_tpm %>%
    inner_join(ensg_hgnc) %>%
    filter(!is.na(hgnc_symbol)) %>%
    select(-gene_id) %>%
    group_by(hgnc_symbol) %>%
    summarise_all(sum)
})

racle = new.env()
racle$expr = tmp_env$racle_expression
racle$expr_mat = racle$expr %>% as.data.frame() %>% column_to_rownames("hgnc_symbol") %>% as.matrix()
racle$ref = tmp_env$racle_cyto %>% 
  rename(sample=`donor ID`) %>% 
  gather(method_cell_type, true_fraction, -sample) %>% 
  mutate(true_fraction=true_fraction/100) %>% 
  inner_join(immunedeconv::cell_type_mapping %>% filter(method_dataset == "racle"))

```

### helper functions for generating bulk samples
```{r}
#' Make a cell fraction vector from single cell data with one specific cell type
#' 
#' @param input_cell_type cell_type for the immune cell to use
#' @param n number of immune cells to use in the sample
#' @param background named vector giving cell counts of background cells. use `NULL` to exclude background cells. 
make_cell_fraction_vector = function(input_cell_type, n, background=n_background_cells) {
  if(input_cell_type == "T cell CD4+") {
    # combine regulatory and non-regulatory CD4+ T cells
    cell_type_stats = pData(single_cell_schelker$eset) %>% 
      group_by(cell_type) %>% 
      count() %>% 
      as.data.frame() %>% 
      column_to_rownames("cell_type")
    ratio = cell_type_stats["T cell regulatory (Tregs)", "n"]/cell_type_stats["T cell CD4+ (non-regulatory)", "n"]
    cell_n =  c(background, round(n*(1-ratio)), round(n*ratio))
    names(cell_n) = c(names(background), "T cell CD4+ (non-regulatory)", "T cell regulatory (Tregs)")
  } else {
    cell_n =  c(background, n)
    names(cell_n) = c(names(background), input_cell_type)
  }
  cell_n
}

background_cells = c(cell_types$cancer, cell_types$other)
n_background_cells = single_cell_schelker$eset %>% pData() %>%
  filter(cell_type %in% background_cells) %>%
  group_by(cell_type) %>%
  count() %>% 
  {'names<-'(.$n, .$cell_type)}

```
