# Mixing study

In this chapter, we will use single cell data from [@Schelker2017] to create simulated bulk RNAseq samples
of which we know the true cell proportions. We futher use these samples to benchmark the performance of different recently published immune deconvolution tools.

## Average fraction of tumour cells
To obtain representatitive simulated samples, we are interested in the average fraction of tumour cells vs immune cells in a mixture.

```{r cancer_cell_type_fractions, results='hold', echo=FALSE}
cell_type_table = pData(single_cell_schelker$eset) %>%
  filter(source %in% c("melanoma", "ascites")) %>%
  mutate(sample=paste(source, donor, sep="_")) %>%
  group_by(source, sample, cell_type) %>%
  summarise(n=n()) %>%
  mutate(freq=n/sum(n))

ggplot(cell_type_table, aes(x=sample, y=freq)) +
  geom_bar(aes(fill=cell_type), stat="identity") +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values=color_scales$cell_types)
```


```{r, echo=FALSE}
cancer_cells = cell_type_table %>%
  filter(cell_type %in% cell_types$cancer)
```

```{r}
cancer_cell_param = MASS::fitdistr(cancer_cells$freq, "normal")
```

The mean cancer fraction is `r round(cancer_cell_param$estimate[1], 2)` ± `r round(cancer_cell_param$estimate[2], 2)`.


## Create simulated bulk tissues
The fractions of a sample are randomly assigned in the following procedure:

1. Draw a random tumour cell content from the distribution fitted above
2. The first half of the samples will use melanoma cells, the second half ovarian cancer cells.
3. Assign the remaining fraction (=not cancer cells)  randomly to the remaining cell types (`r cell_types$available`)

```{r mixing_make_proportions, echo=FALSE}
set.seed(42)

is_melanoma = c(rep(TRUE, 50), rep(FALSE, 50))

cell_fractions = lapply(is_melanoma, function(melanoma) {
  cancer_fraction = rnorm(1, mean=cancer_cell_param$estimate[1], sd=cancer_cell_param$estimate[2])
  cancer_fraction = ifelse(cancer_fraction < 0, 0, cancer_fraction)
  cancer_fraction = ifelse(cancer_fraction > 1, 1, cancer_fraction)

  # associate fraction randomly ot Melanoma or Ovarian
  if(melanoma) {
    cancer_fractions = list(cancer_fraction, 0)
  } else {
    cancer_fractions = list(0, cancer_fraction)
  }
  names(cancer_fractions) = cell_types$cancer

  # compute random fractions for other cells
  use_cell_types = c(cell_types$immune_cells, cell_types$other)
  remaining_fraction = 1 - cancer_fraction
  rnd = sample(0:100, length(use_cell_types), TRUE)
  tmp_fractions = (remaining_fraction * rnd) / sum(rnd)
  names(tmp_fractions) = use_cell_types

  c(cancer_fractions, tmp_fractions)
}) %>% bind_rows()

testthat::test_that("all fractions sum up to 1", expect_equal(apply(cell_fractions, 1, sum), rep(1, 100)))
```

Here, we generate a simulated bulkRNA Expressionset:
```{r, cache=TRUE}
set.seed(42)
bulk_eset = make_bulk_eset(eset=single_cell_schelker$eset,
                           cell_fractions = cell_fractions,
                           n_cells=500)
```


## Run the deconvolution
We first run all methods that are integrated in the `immune_deconvolution_methods` are package automatically in a loop.
```{r mixing_deconvolution, cache=TRUE, message=FALSE, echo=FALSE, warning=FALSE, results='hide'}
timer_indications = ifelse(is_melanoma, "SKCM", "OV")
all_results = foreach(method=immunedeconv::deconvolution_methods,
                      .final = function(x) {setNames(x, immunedeconv::deconvolution_methods)}) %do% {
  deconvolute(bulk_eset, method, column="gene_symbol", indications = timer_indications)
}
```



## Results

```{r, include=FALSE}
res_mixing_study = new.env()
```

The following methods have absolute (EPIC, quanTIseq) or pseudo-absolute (CIBERSORT abs. mode, xCell) scores, and we test how well they perform
in terms of absolute deviation
```{r}
abs_methods = c("cibersort_abs", "epic", "quantiseq", "xcell")
```

We perform an evaluation for the following cell types.
Note that some cell types are redundant (T cell CD4+ is a super-category of Tregs and non-regulatory CD4+ T cells).
As some methods provide deconvolution only at the CD4+ level, we compare both categories:
```{r}
show_cell_types = c("B cell", "Dendritic cell", "Macrophage/Monocyte",
                    "NK cell", "T cell CD4+", "T cell CD4+ (non-regulatory)",
                    "T cell regulatory (Tregs)", "T cell CD8+",
                    "Cancer associated fibroblast", "Endothelial cell")
```

The following methods do not have signatures to quantify "Macrophages/Monocytes", but only a Macrophage signature. Unfortunately,
our single cell dataset does not distinguish between Macrophages and Monocytes. As we are only comparing correlations here, we still
benchmark the Macrophage signature on the "Macrophage/Monocyte" data, as an increase in both should also lead to an increase in Macrophages only.
To be fair, we label the results accordingly.

```{r}
macrophage_signature_only = c("epic", "timer")
```

Here, we map the results back to the "gold standard". We aggregate the results of the different methods
into a single table and clean it up for further processing.
```{r mixing_data_postprocessing, cache=TRUE, echo=FALSE}
all_results2 = lapply(names(all_results), function(method) {
  map_result_to_celltypes(all_results[[method]], show_cell_types, method) %>%
    as_tibble(rownames="cell_type") %>%
    mutate(method=method)
})

all_results_tidy = all_results2 %>%
  bind_rows() %>%
  na.omit() %>%
  gather(sample, estimate, -cell_type, -method) %>%
  mutate(macrophage_only=if_else(method %in% macrophage_signature_only & cell_type == "Macrophage/Monocyte", "yes", "no")) %>%
  arrange(method, cell_type, sample)

gold_standard = pData(bulk_eset) %>%
  t() %>%
  as_tibble(rownames="cell_type") %>%
  map_result_to_celltypes(show_cell_types, "schelker_single_cell") %>%
  rownames_to_column("cell_type") %>%
  gather(sample, true_fraction, -cell_type)
```

```{r, cache=TRUE}
results_with_gold_standard = inner_join(all_results_tidy, gold_standard, by=c("sample", "cell_type"))
```

### Correlation plots
```{r mixing_correlation_plot, fig.width=18, fig.height=10, echo=FALSE}
results_with_gold_standard %>%
  ggplot(aes(x=true_fraction, y=estimate)) +
  facet_grid(method ~ cell_type, scales="free") +
  geom_point() +
  stat_smooth(aes(color=method), method="lm") +
  scale_color_manual(values=color_scales$methods) +
  scale_x_continuous(breaks=c(.2, .3)) +
  stat_cor()
```


### Calculate correlations for each method and cell type
```{r mixing_correlation_summary, fig.width=8, fig.height=6, echo=FALSE}
make_cor = function(estimate, true_fraction) {
  test_res = cor.test(estimate, true_fraction, method="pearson")
  data.frame(
    pearson=test_res$estimate,
    conf_int_lower=test_res$conf.int[1],
    conf_int_upper=test_res$conf.int[2]
  )
}

correlations = results_with_gold_standard %>%
  group_by(cell_type, method, macrophage_only) %>%
  do(make_cor(.$estimate, .$true_fraction)) %>% 
  ungroup()

res_mixing_study$correlations = correlations
```

```{r export_correlations}
correlations_tab = correlations %>% select(cell_type, method, pearson) %>% 
  spread(cell_type, pearson)

write_tsv(correlations_tab, "../results/tables/mixing_study_correlations.tsv", na="")

```


```{r, fig.width=8, fig.height=6, echo=FALSE}
correlations %>%
  ggplot(aes(x=method, y=pearson, alpha=macrophage_only)) +
    geom_bar(aes(fill=method), stat="identity") +
    geom_errorbar(aes(ymin=conf_int_lower, ymax=conf_int_upper), width=.2) +
    facet_wrap(~cell_type, nrow=2, labeller = label_wrap_gen()) +
    theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position="top",
          strip.text.x = element_text(size=9)) +
    scale_fill_manual(values=color_scales$methods) +
    scale_alpha_manual(values=c("yes"=.3, "no"=1.))
```


### Deviation from slope (x=y) for all absolute methods
If the absolute devoncolution was perfect, the slope would equal 1 ($x \mapsto y$ ). 
The figure shows the deviation from the 1, i.e. values > 0 indicate the method over-estimates the fraction
of these cells; values < 0 indicate the method under-estiamtes the fraction of these cells. 

Confidence intervals are two-sided 95% confidence intervales derived using `confint` and `lm`. 
```{r mixing_abs, echo=FALSE, cache=TRUE}
get_slope = function(df) {
  fit = lm(estimate~true_fraction, data=df)
  conf_int = confint(fit, "true_fraction", level=0.95)
  data.frame(
    slope = fit$coefficients[2],
    ci_lower = conf_int[1],
    ci_upper = conf_int[2]
  )
}


deviation_from_slope = results_with_gold_standard %>%
  filter(method %in% abs_methods) %>% 
  group_by(macrophage_only, cell_type, method) %>% 
  do(get_slope(.)) %>% 
  mutate(deviation=slope-1, ci_lower=ci_lower-1, ci_upper=ci_upper-1)

res_mixing_study$deviation_from_slope = deviation_from_slope
```

```{r, fig.width=8, fig.height=4, echo=FALSE}
deviation_from_slope %>% 
 ggplot(aes(x=method, y=deviation, alpha=macrophage_only)) +
    geom_bar(aes(fill=method), stat="identity") +
    geom_errorbar(aes(ymin=ci_lower, ymax=ci_upper), width=.2) +
    facet_wrap(~cell_type, nrow=2, labeller = label_wrap_gen()) +
    theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position="top",
          strip.text.x = element_text(size=9)) +
    scale_fill_manual(values=color_scales$methods) +
    scale_alpha_manual(values=c("yes"=.3, "no"=1.)) +
    geom_hline(yintercept=0, col="grey")
```

