---
title: "R Notebook"
output: html_notebook
---

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr,
               tidyverse,
               cowplot,
               Biobase,
               pheatmap,
               magrittr,
               assertthat,
               testthat)
```

Load dataset from schelker et al. 
```{r}
load("../data/single_cell_schelker.rda")
```

Let's repreduce the tsne plot from the publication
```{r, fig.width=10, fig.height=10}
pData(single_cell_schelker) %>% 
  ggplot(aes(x=tsneX1, y=tsneX2, colour=cell_type)) +
           geom_point(size=1) + 
           theme(legend.position="bottom")
```

Summary of the cell types
```{r}
pData(single_cell_schelker) %>% group_by(cell_type) %>% count() %>% knitr::kable()
```

Remove cells of Unknown type for downstream analysis
```{r}
valid_cells = which(pData(single_cell_schelker)$cell_type!="Unknown")

single_cell_schelker2 = single_cell_schelker[,valid_cells]

assert_that(!any(pData(single_cell_schelker2)$cell_type == "Unknown"))

available_cell_types = pData(single_cell_schelker2) %>% select(cell_type) %>% distinct() %>% pull(cell_type)

```


To obtain representatitive simulated samples, we are interested in the average fraction of tumour cells vs immune cells in a mixture. 

```{r}
cell_type_table = pData(single_cell_schelker2) %>%
  filter(source %in% c("melanoma", "ascites")) %>%
  mutate(sample=paste(source, donor, sep="_")) %>% 
  group_by(source, sample, cell_type) %>% 
  summarise(n=n()) %>%
  mutate(freq=n/sum(n))

ggplot(cell_type_table, aes(x=sample, y=freq)) +
  geom_bar(aes(fill=cell_type), stat="identity") + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1))


cancer_cells = cell_type_table %>% 
  filter(cell_type %in% c("Melanoma cells", "Ovarian carcinoma cells")) 


cancer_cell_param = MASS::fitdistr(cancer_cells$freq, "normal")
```



```{r}
devtools::load_all("/storage/home/sturm/projects/immune_deconvolution_methods")
```

```{r}
set.seed(42)

cancer_cells = c("Melanoma cells", "Ovarian carcinoma cells")
immune_cells = available_cell_types[!(available_cell_types %in% cancer_cells)]
is_melanoma = c(rep(TRUE, 50), rep(FALSE, 50))

cell_fractions = lapply(is_melanoma, function(melanoma) {
  cancer_fraction = rnorm(1, mean=cancer_cell_param$estimate[1], sd=cancer_cell_param$estimate[2])
  cancer_fraction = ifelse(cancer_fraction < 0, 0, cancer_fraction)
  cancer_fraction = ifelse(cancer_fraction > 1, 1, cancer_fraction)
  
  # associate fraction randomly ot Melanoma or Ovarian
  if(melanoma) {
    cancer_fractions = list(cancer_fraction, 0)
  } else {
    cancer_fractions = list(0, cancer_fraction)
  }
  names(cancer_fractions) = cancer_cells
  
  # compute random fractions for other cells
  remaining_fraction = 1 - cancer_fraction
  rnd = sample(0:100, length(immune_cells), TRUE)
  tmp_fractions = (remaining_fraction * rnd) / sum(rnd)
  names(tmp_fractions) = immune_cells
  
  c(cancer_fractions, tmp_fractions)
}) %>% bind_rows()

testthat::test_that("all fractions sum up to 1", expect_equal(apply(cell_fractions, 1, sum), rep(1, 100)))
```


```{r}
set.seed(42)
bulk_eset = make_bulk_eset(eset=single_cell_schelker2, cell_fractions = cell_fractions, n_cells=500)
```


## run all methods and aggreate the results in a single table
```{r, fig.width=15, fig.height=5}
all_results = lapply(immunedeconv::deconvolution_methods, function(method) {
  deconvolute(bulk_eset, method, column="gene_symbol") %>%
    mutate(method=method)
})
```

# run methods that require special "treatment"
```{r}
eset_mat = eset_to_matrix(bulk_eset, "gene_symbol")
res_cibersort = immunedeconv::deconvolute_cibersort(eset_mat, 
                                                    "../lib/CIBERSORT/CIBERSORT.R",
                                                    "../lib/CIBERSORT/LM22.txt") 
res_cibersort %<>% mutate(method="cibersort")

res_cibersort_abs = immunedeconv::deconvolute_cibersort(eset_mat, 
                                                    "../lib/CIBERSORT/CIBERSORT.R",
                                                    "../lib/CIBERSORT/LM22.txt",
                                                    absolute = TRUE) 
res_cibersort_abs %<>% mutate(method="cibersort_abs")
```

```{r}
eset_mat = eset_to_matrix(bulk_eset, "gene_symbol")
immunedeconv::export_for_timer(eset_mat[,is_melanoma], path="../results/timer_input_melanoma.tsv")
immunedeconv::export_for_timer(eset_mat[,!is_melanoma], path="../results/timer_input_ovarian.tsv")

res_timer_melanoma = immunedeconv::import_from_timer("../data/timer/timer_result_melanoma.csv")
res_timer_ovarian = immunedeconv::import_from_timer("../data/timer/timer_result_ovarian.csv")
res_timer = inner_join(res_timer_melanoma, res_timer_ovarian, by="cell_type") %>%
  mutate(method="timer")
```



```{r}
all_results %<>% 
  c(list(res_cibersort)) %>%
  c(list(res_cibersort_abs)) %>% 
  c(list(res_timer))

# summarise cell_types from the method that only map to a single cell_type in the reference dataset (namely monocytes and macrophages)
all_results2 = lapply(all_results, function(res) {
  res %>% 
    map_results_to_dataset("single_cell_schelker") %>% 
    group_by(method, single_cell_schelker) %>%   
    summarise_all(funs(sum)) 
})

all_results_tidy = all_results2 %>%
  bind_rows() %>%
  rename(cell_type = single_cell_schelker) %>%
  gather(sample, estimate, -cell_type, -method) %>%
  arrange(method, cell_type, sample)

gold_standard = pData(bulk_eset) %>%
  rownames_to_column("sample") %>%
  gather(cell_type, true_fraction, -sample)

results_with_gold_standard = all_results_tidy %>%
  inner_join(gold_standard, by=c("sample", "cell_type"))

```

## correlation plot
```{r, fig.width=12, fig.height=8}
results_with_gold_standard %>%
  ggplot(aes(x=true_fraction, y=estimate)) +
  facet_grid(method ~ cell_type, scales="free_y") +
  geom_point() +
  stat_smooth(method="lm")
```


## Calculate correlations for each method and cell type
```{r}
correlations = results_with_gold_standard %>%
  group_by(cell_type, method) %>%
  summarise(pearson=cor(estimate, true_fraction, method="pearson"))

correlations %>% ggplot(aes(x=method, y=pearson)) +
  geom_bar(stat="identity") +
  facet_wrap(~cell_type) + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1))
```



