# Validation with real data
```{r}
res_validation = new.env()

```
We the curated datasets with gold standard as an additional validation on top of the mixing study. 

We use the following three validation datasets (see table \@ref(tab:validation_data)), which are the only ones
in the tables for which we could obtain both gene expression and validation data
```{r, cache=TRUE}
datasets = list(
  racle=racle,
  hoeck=hoeck,
  schelker_ovarian=schelker_ovarian
)
```

We use the following cell types, which are available in (some of) the datasets
```{r}
use_cell_types = c("T cell", "T cell CD8+", "T cell CD4+",
                   "Macrophage/Monocyte", "B cell", 
                   "Dendritic cell", "NK cell", "Neutrophil")
```

```{r, echo=FALSE, cache=TRUE, message=FALSE, echo=FALSE, warning=FALSE, results='hide'}
# run deconvolution and include TIMER results

## process a deconvolution result
process_result = function(result, method, dataset_name) {
  result %>% 
    map_result_to_celltypes(use_cell_types, method) %>% 
    as_tibble(rownames="cell_type") %>% 
    na.omit() %>% 
    gather(sample, estimate, -cell_type) %>% 
    mutate(method=method, dataset=dataset_name)
}

timer_cancer_type = list(racle="SKCM", # actually PBMC
                         hoeck="SKCM", 
                         schelker_ovarian="OV")


# Run the deconvolution...
all_results = foreach(dataset=datasets, dataset_name=names(datasets), .combine=bind_rows) %:%
  foreach(method = immunedeconv::deconvolution_methods, .combine=bind_rows) %do% {
    timer_indications = rep(timer_cancer_type[[dataset_name]], ncol(dataset$expr_mat))
    deconvolute(dataset$expr_mat, method, indications=timer_indications) %>%
      process_result(method, dataset_name)
}

all_refs = foreach(dataset=datasets, dataset_name=names(datasets), .combine=bind_rows) %do% {
  dataset$ref %>% 
    select(sample, cell_type, true_fraction) %>%
    spread(sample, true_fraction) %>% 
    map_result_to_celltypes(use_cell_types) %>% 
    as_tibble(rownames="cell_type") %>% 
    gather(sample, true_fraction, -cell_type) %>% 
    mutate(dataset=dataset_name) %>% 
    na.omit()
}
```

Here, we combine the predictions with the 'gold standard' reference data. 
```{r, cache=TRUE}
all_results_ref = inner_join(all_results, all_refs, 
                             by=c("sample"="sample", "dataset"="dataset", "cell_type"="cell_type"))

res_validation$all_results = all_results_ref
```


```{r, fig.width=10, fig.height=16, echo=FALSE}
all_results_ref %>%
  filter(!(cell_type == "T cell" & dataset != "hoeck")) %>% 
  ggplot(aes(x=true_fraction, y=estimate)) +
    geom_point(aes(color=cell_type)) + 
    scale_color_manual(values=color_scales$validation) + 
    facet_grid(method~dataset, scales = "free_y") + 
    theme_bw() + 
    theme(legend.position = "top",
          strip.text.x = element_text(size=9)) + 
    background_grid(major="xy") + 
    stat_cor() 
```

Observation: all methods work better on PBMC than they do on cancer cells. 
