# Detection limit and false positive predictions
```{r, include=FALSE}
res_sensitivity = new.env()
```

How many cells of a certain type do we need for a method to detect immune cell infiltration (=Detection limit)?
How many cells of a certain type are detected, although we know they are not there (=false positives)?

We again use the single cell dataset to simulate samples that consist of background cells
(i.e. non-immune cells: fibroblasts, endothelial cells cancer cells) and add an increasing amount
of immune cells of a certain type.

We define the detection limit as the minimal fraction at which the method the abundance of the cell type to
be significantly different from zero. We define false positives as the predicted fraction of a certain cell type
at zero inflitration level.

For each amount of immune cells, we generate 5 random samples to compute a confidence interval.

We use the following cell types:
```{r}
show_cell_types = c("B cell", "Dendritic cell", "Macrophage/Monocyte",
                    "NK cell", "T cell CD4+", "T cell CD8+", 
                    "T cell CD4+ (non-regulatory)", "T cell regulatory (Tregs)", 
                    "Cancer associated fibroblast", "Endothelial cell")
cancer_background = c("Melanoma cell"=700, "Ovarian carcinoma cell"=300)
N_BACKGROUND = 1000
```

```{r, echo=FALSE}
# watch out when changing order! we rely on the order to execute
# the samples in batches. 
N_BATCH = 5
n_immune_cells = rep(c(seq(0, 50, 5)
                     , seq(60, 100, 10)
                     , seq(120, 300, 20)
                     , seq(350, 500, 50)
                     ), N_BATCH)
```


Generate dataset with only cancer cells as background: 
```{r sens_generate_data, cache=FALSE, echo=FALSE}
sensitivity_analysis_file = "../results/cache/sensitivity_analysis_dataset.rda"
if(!file.exists(sensitivity_analysis_file)) {
  set.seed(42)
  res = foreach(input_cell_type = show_cell_types,
                .final = function(x) setNames(x, show_cell_types)) %dopar% {
                    background=cancer_background
                    assert(sum(background) == N_BACKGROUND)
                    expr_mat = lapply(n_immune_cells, function(k) {
                      cell_n = make_cell_fraction_vector(input_cell_type, k, background)
                      cell_frac = cell_n/sum(cell_n)
                      make_random_bulk(eset=single_cell_schelker$eset, cell_fractions = cell_frac, n_cells = sum(cell_n))
                    }) %>% bind_cols() %>% as.matrix()

                    rownames(expr_mat) = rownames(single_cell_schelker$eset %>% exprs())
                    expr_mat
    }

  sensitivity_analysis_dataset = res
  save(sensitivity_analysis_dataset, file=sensitivity_analysis_file)
} else {
  load(sensitivity_analysis_file)
}
```

Generate dataset with cancer and all other immune cells as background: 
```{r sens_generate_data_immune_bg, cache=FALSE, echo=FALSE}
sensitivity_analysis_file = "../results/cache/sensitivity_analysis_dataset_immune_background.rda"
if(!file.exists(sensitivity_analysis_file)) {
  set.seed(42)
  res = foreach(input_cell_type = show_cell_types,
                .final = function(x) setNames(x, show_cell_types)) %dopar% {
                    background_cts = c(cell_types$immune_cells, cell_types$other, cell_types$cancer)
                    # for CD4+, which is a supercategory, exclude both child cell types. 
                    if(input_cell_type == "T cell CD4+") {
                      exclude_cell_types = c("T cell regulatory (Tregs)", "T cell CD4+ (non-regulatory)") 
                    } else {
                      exclude_cell_types = input_cell_type
                    }
                    background_cts = background_cts[!background_cts %in% exclude_cell_types]
                    rnd = sample(0:100, length(background_cts), TRUE)
                    background = round(rnd / sum(rnd) * 1000, 0)
                    names(background) = background_cts
                    assert("1000 cells in background", abs(sum(background) - 1000) <= length(background))
                    expr_mat = lapply(n_immune_cells, function(k) {
                      cell_n = make_cell_fraction_vector(input_cell_type, k, background)
                      cell_frac = cell_n/sum(cell_n)
                      make_random_bulk(eset=single_cell_schelker$eset, cell_fractions = cell_frac, n_cells = sum(cell_n))
                    }) %>% bind_cols() %>% as.matrix()

                    rownames(expr_mat) = rownames(single_cell_schelker$eset %>% exprs())
                    expr_mat
    }

  sensitivity_analysis_dataset_immune_bg = res
  save(sensitivity_analysis_dataset_immune_bg, file=sensitivity_analysis_file)
} else {
  load(sensitivity_analysis_file)
}
```

```{r sens_deconvolute, fig.height=18, fig.width=25, cache=TRUE, echo=FALSE, message=FALSE}
deconvolute_dataset = function(dataset, n_immune_cells) {
  tmp_cell_types = names(dataset)
  expr_mat = do.call(cbind, dataset)
  
  input_cell_types = unlist(lapply(tmp_cell_types, function(cell_type) {
    rep(cell_type, length(n_immune_cells))
  }))
  colnames(expr_mat) = unlist(lapply(tmp_cell_types, function(cell_type) {
    paste0(cell_type, "_", seq(1, length(n_immune_cells)))
  }))
  
  res = foreach(
    method = config$deconvolution_methods,
    .final = function(x)
      setNames(x, config$deconvolution_methods)
  ) %:%
    foreach(i = 1:N_BATCH,
            .combine = full_join) %dopar% {
              batch_length = length(unique(n_immune_cells))   # number of samplese per batch
              n_samples = length(n_immune_cells)              # number of samples per cell type
              ind = ((i - 1) * batch_length + 1):(i * batch_length) # indices for first cell type
              ind_ct = lapply(seq(0, length(tmp_cell_types) - 1), function(k) {
                (k * n_samples) + ind
              }) %>% unlist()                                 # indices for all cell types
              # use SKCM, although samples will also contain ovarian cancer acsites cells.
              timer_indications = rep("SKCM", ncol(expr_mat))
              deconvolute(expr_mat[,ind_ct], method, indications = timer_indications[ind_ct])
            }
  
  res

}

sensitivity_analysis_res = deconvolute_dataset(sensitivity_analysis_dataset, n_immune_cells)
sensitivity_analysis_res_immune_background = deconvolute_dataset(sensitivity_analysis_dataset_immune_bg, n_immune_cells)

save(sensitivity_analysis_res, sensitivity_analysis_res_immune_background, file = "../results/cache/sensitivity_analysis_res.rda")
```

```{r sens_process_data, cache=TRUE, echo=FALSE, message=FALSE}
# post process the deconvolution results
post_process_deconvolution_results = function(result) {
 all_results = foreach(method = config$deconvolution_methods,
                      .combine=bind_rows) %do% {
                        tmp_res = result[[method]] %>%
                          map_result_to_celltypes(show_cell_types, method) %>%
                          na.omit()
                       
                        cell_type_sample_id = str_split_fixed(colnames(tmp_res), "_", 2)
                        n_cell_df = tibble(sample=colnames(tmp_res),
                                           input_cell_type = cell_type_sample_id[,1],
                                           n_immune_cells = n_immune_cells[as.numeric(cell_type_sample_id[,2])],
                                           n_background_cells = N_BACKGROUND)
                        
                        tmp_res %>%
                          as_tibble(rownames = "cell_type") %>%
                          gather(sample, estimate, -cell_type) %>%
                          mutate(method=method) %>%
                          inner_join(n_cell_df, by="sample")
                      } %>%
  mutate(frac_immune_cells = n_immune_cells/N_BACKGROUND)
 
  all_results
}

compute_conf_int = function(all_results) {
 all_results %>%
  select(-sample) %>%
  group_by_at(vars(-estimate)) %>%
  summarise_all(funs(mean, sd, n=length)) %>%
  mutate(ci=qt(0.975,df=n-1)*sd/sqrt(n)) %>%
  ungroup() 
}

all_results = list(
  "cancer_bg" = post_process_deconvolution_results(sensitivity_analysis_res),
  "immune_bg" = post_process_deconvolution_results(sensitivity_analysis_res_immune_background)
)

all_results_ci = lapply(all_results, compute_conf_int)
```

## Predictions with increasing immune cell content
```{r, fig.height=18, fig.width=32, echo=FALSE, fig.cap="We add an increasing amount of each cell type independently to the 1,800 background cells. The figure shows the predictions for each cell type and method at increasing infiltration levels. Tip: right-click, 'open in new tab' to enlarge!"}
all_results_ci$cancer_bg %>%
  ggplot(aes(x=frac_immune_cells, y=mean, colour=cell_type)) +
  #   geom_line() +
    geom_errorbar(aes(ymin=mean-ci, ymax=mean+ci)) +
    theme_bw() +
    theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position="top") +
    facet_grid(method~input_cell_type, scales = "free_y") +
    scale_color_manual(values = color_scales$immune_cells) +
    ggtitle("cancer background")

all_results_ci$immune_bg %>%
  ggplot(aes(x=frac_immune_cells, y=mean, colour=cell_type)) +
  #   geom_line() +
    geom_errorbar(aes(ymin=mean-ci, ymax=mean+ci)) +
    theme_bw() +
    theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position="top") +
    facet_grid(method~input_cell_type, scales = "free_y") +
    scale_color_manual(values = color_scales$immune_cells) + 
    ggtitle("immune background")
```


## Detection Limit and False positives
```{r detection-limit, include=FALSE}
t_test = function(x, y) {
  res = t.test(x, y, paired=FALSE, alternative = "greater")
  res$p.value
}

make_prediction_at_zero_data = function(all_results) {
  all_results %>% filter(frac_immune_cells == 0) %>%
    filter(cell_type == input_cell_type) %>% 
    select(method, cell_type, background=estimate) %>% 
    group_by(method, cell_type) %>% 
    mutate(sample_n = 1:n()) %>% 
    arrange(method, cell_type)
}

make_background_prediction_data = function(all_results, prediction_at_zero) {
  background_data = all_results %>%
  filter(cell_type == input_cell_type) %>% 
    select(method, cell_type, estimate, frac_immune_cells) %>% 
    group_by(method, cell_type, frac_immune_cells) %>% 
    mutate(sample_n = 1:n()) %>% 
    arrange(method, cell_type, frac_immune_cells) %>% 
    inner_join(prediction_at_zero) %>% 
    group_by(method, cell_type, frac_immune_cells) %>%
    summarise(n=n(), mean=mean(estimate), sd=sd(estimate), p_value=t_test(estimate, background)) %>%
    mutate(ci=qt(0.975,df=n-1)*sd/sqrt(n))
  
  background_data
}

make_detection_limit_data = function(background_data) {
  background_data %>% 
  group_by(method, cell_type) %>% 
  # we look for the minimal fraction, at which the prediction is significantly different from the background. 
  filter(p_value < 0.05) %>% 
  summarise(min_frac = min(frac_immune_cells))
}

make_false_positive_data = function(prediction_at_zero) {
  prediction_at_zero %>% 
    group_by(method, cell_type) %>% 
    summarise(n=n(), mean = mean(background), sd=sd(background)) %>%
    mutate(ci=qt(0.975,df=n-1)*sd/sqrt(n))
}

prediction_at_zero = lapply(all_results, make_prediction_at_zero_data)
background_data = mapply(make_background_prediction_data, all_results, prediction_at_zero, SIMPLIFY=FALSE) 
data_detection_limit = lapply(background_data, make_detection_limit_data)
data_false_positives = lapply(prediction_at_zero, make_false_positive_data)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, warning=FALSE, fig.width=12, fig.height=9, fig.cap="Detection limit and false positive predictions. The plots show the average estimate of each method for five simulated bulk RNA-seq datasets randomly sampled from ~1,800 non-immune cells and an increasing fraction of a specific immune cell type. The grey ribbon indicates the standard deviation. The red line refers to the detection limit, i.e. the minimal fraction of an immune cell type needed for a method to reliably detect its abundance as significantly different from zero (p-value < 0.05, two-sided t-test). The blue line refers to the false-positive prediction rate, i.e. the average estimate of a method with no immune cells present. "}


fun_breaks2 = function(limits) {
  breaks = signif(max(limits) * c(0, 0.5),1)
  names(breaks) = attr(breaks, "labels")
  breaks
}

bind_dataframes = function(data_list, name_column="dataset") {
  lapply(names(data_list), function(i) {
    data_list[[i]] %>% mutate(!!name_column := i)
  }) %>% bind_rows()
}

background_data %>%
  bind_dataframes() %>% 
  ggplot(aes(x=frac_immune_cells, y=mean)) + 
    geom_ribbon(aes(ymin=mean-ci, ymax=mean+ci, fill=dataset), alpha=.2) + 
    geom_point(aes(colour=dataset), size=.2, shape=4) + 
    # geom_errorbar(aes(ymin=mean-ci, ymax=mean+ci)) + 
    panel_border() + 
    facet_grid(method ~ cell_type, scales = "free_y") + 
    scale_x_continuous(breaks=c(.0, .3)) +
    scale_y_continuous(breaks=fun_breaks2) +
    geom_hline(mapping=aes(yintercept=0, colour="zero"), linetype="dashed", show.legend=FALSE) + 
    geom_vline(mapping=aes(xintercept = 0, color="zero"), linetype="dashed", show.legend=FALSE) + 
    # geom_vline(data=data_detection_limit$cancer_bg, mapping=aes(xintercept=min_frac, colour="detection limit"), show.legend = FALSE) + 
    geom_errorbar(data=bind_dataframes(data_false_positives), mapping=aes(ymin=mean-ci, ymax=mean+ci, x=-0.03, colour=dataset), width=0.04, position="dodge") + 
    # geom_crossbar(data=bind_dataframes(data_false_positives), mapping=aes(y=mean, x=0, ymin=mean, ymax=mean, colour=dataset), width=0.05, position="dodge") + 
    geom_hline(yintercept = 0, mapping = aes(colour="detection limit"), alpha=0) + # fake line to fix legend
    ylab("average estimate") + 
    xlab("fraction of spike-in cells") + 
    theme(legend.position = "top", strip.text=element_text(size=8), 
          panel.spacing = unit(1, "mm"), axis.text = element_text(size=10)) + 
    scale_color_manual(values=c("zero"="grey", "detection limit"="red", "false positive fraction"="blue", "cancer_bg"="#1f78b4", "immune_bg"="#33a02c"), guide=guide_legend("performance measure", override.aes = list(alpha=1)))  
```

```{r, include=FALSE}
res_sensitivity$all_results = all_results
res_sensitivity$background_data = background_data
res_sensitivity$data_detection_limit = data_detection_limit
res_sensitivity$data_false_positives = data_false_positives

```


```{r, sens-spec-plot, fig.height=5, fig.width=12, echo=FALSE, fig.cap="Predicted amount of a certain cell type while it is actually absent. This analysis is based on all of the data above where only background cells are present, i.e. fraction of immune cells = 0."}
prediction_at_zero %>%
  bind_dataframes() %>% 
  filter(method != "mcp_counter") %>% 
  ggplot(aes(x=cell_type, y=background, colour=dataset)) +
    geom_boxplot(position="dodge") +
    facet_wrap(~cell_type, nrow=1) 
    # theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1),
    #       legend.position="top",
    #       strip.text.x = element_text(size=9)) 
```

