# Spillover Analysis
```{r, include=FALSE}
res_spillover = new.env()
```

In this chapter, we investigate which other cell types a method predicts,
if there is actually only a certain cell type present. In FACS, this phenomenon is
known as "spillover".

To this end, we use three datasets

 * *immune reference*: bulk RNA seq profiles from sorted immune cells (=quanTIseq training data)
 * *artificial bulk*: simulated bulk RNA seq profiles from single cells (e.g. only T cells)
 * *artificial bulk with background*: simulated bulk RNA seq profiles from single cells with ~80% other cells (cancer, fibroblasts, ...)

We test the following cell types:
```{r}
show_cell_types = c("B cell", "Dendritic cell", "Macrophage/Monocyte",
                    "NK cell", "T cell CD4+", "T cell CD8+")
```


```{r, cache=TRUE, include=FALSE}
# Create datasets

set.seed(42)

sample_types = rep(show_cell_types, 5)
## make artificial bulk sample
artificial_bulk = lapply(sample_types, function(cell_type) {
  cell_n = make_cell_fraction_vector(cell_type, k=500, background=NULL)
  cell_frac = cell_n/sum(cell_n)
  make_random_bulk(single_cell_schelker$eset, cell_frac, n_cells=500)
}) %>% bind_cols() %>% as.matrix()

rownames(artificial_bulk) = rownames(single_cell_schelker$eset %>% exprs())


## make artificial bulk samples with background
artificial_bulk_bg = lapply(sample_types, function(cell_type) {
  cell_n = make_cell_fraction_vector(cell_type, k=100,
                                     background=round(400*(cell_types$n_background_cells/sum(cell_types$n_background_cells))))
  assert("sample has 500 cells", sum(cell_n) >= 499 & sum(cell_n) <= 501)
  cell_frac = cell_n/sum(cell_n)
  make_random_bulk(single_cell_schelker$eset, cell_frac, n_cells=500)
}) %>% bind_cols() %>% as.matrix()

rownames(artificial_bulk_bg) = rownames(single_cell_schelker$eset %>% exprs())
```


```{r spec_generate_data, cache=TRUE, include=FALSE}
# Run the deconvolution
datasets = list(
  immune_reference = immune_cell_reference$expr_mat,
  artificial_bulk = artificial_bulk,
  artificial_bulk_bg = artificial_bulk_bg
)
dataset_gold_standard = list(
  immune_reference = immune_cell_reference$sample_description %>%
    select(sample, true_cell_type=cell_type),
  artificial_bulk = tibble(sample=colnames(artificial_bulk), true_cell_type=sample_types),
  artificial_bulk_bg = tibble(sample=colnames(artificial_bulk), true_cell_type=sample_types)
)

if(!file.exists(all_results_file)) {
  all_results = foreach(dataset = names(datasets),
                        .final = function(x) setNames(x, names(datasets))) %:%
                  foreach(method = config$deconvolution_methods,
                          .final = function(x) setNames(x, config$deconvolution_methods)) %dopar% {
                          # for timer, we use SKCM, although the samples either
                          # include no tumor cells, or both melanoma and OV cells.
                          timer_indications = rep("SKCM", ncol(datasets[[dataset]]))
                          deconvolute(datasets[[dataset]], method, indications=timer_indications)
                }
  save(all_results, file=all_results_file)
}
load(all_results_file)
```

```{r, cache=TRUE, message=FALSE, echo=FALSE}
# Postprocess the deconvolution results
all_results2 =  foreach(dataset = names(datasets),
                        gold_standard = dataset_gold_standard,
                        .combine = bind_rows) %:%
                  foreach(method = config$deconvolution_methods,
                          .combine = bind_rows) %do% {
                            all_results[[dataset]][[method]] %>%
                              map_result_to_celltypes(show_cell_types, method) %>%
                              as_tibble(rownames = "cell_type") %>%
                              gather(sample, estimate, -cell_type) %>%
                              mutate(method=method, dataset=dataset) %>%
                              inner_join(gold_standard)
                } %>%
  na.omit()

res_spillover$all_results = all_results2
```

## Complete Spillover Matrix
```{r, fig.width=24, fig.height=24, echo=FALSE, fig.cap="This figure shows the spillover for all methods, cell types and datasets"}
all_results2 %>%
  filter(true_cell_type %in% show_cell_types, cell_type %in% show_cell_types) %>%
  ggplot(aes(x=cell_type, y=estimate, fill=dataset)) +
    geom_boxplot(position = position_dodge()) +
    facet_grid(method~true_cell_type+dataset, scales="free_y") +
    theme_bw() +
    theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1)) +
    background_grid(major = "xy", minor = "none")

```


## Summary figure: Signal to noise ratio
Noise is defined as the sum of all predictions of other cell types than the one that is actually present.
Signal is defined as the predicted fraction of the cell type that is actually present.
The signal ratio is defined as $\frac{\text{signal}}{\text{signal+noise}}$. Higher values
indicate less noise (i.e. predictions of cell types that are not there).

```{r, fig.width=10, fig.height=8, echo=FALSE}
signal_noise_res = all_results2 %>%
  mutate(signal_noise=if_else(cell_type == true_cell_type, "signal", "noise")) %>%
  group_by(dataset, sample, method, true_cell_type, signal_noise) %>%
  summarise(estimate=sum(estimate)) %>%
  spread(signal_noise, estimate) %>%
  mutate(noise_ratio = noise/(noise+signal)) %>%
  mutate(signal_ratio = signal/(noise+signal)) %>%
  ungroup() %>%
  na.omit()

res_spillover$signal_noise = signal_noise_res
```

```{r, fig.width=10, fig.height=8, echo=FALSE, fig.cap="Signal ratios for each cell type and dataset. "}
signal_noise_res %>%
  ggplot(aes(x=method, y=signal_ratio, fill=method)) +
  geom_boxplot(position = position_dodge()) +
  facet_grid(dataset~true_cell_type) +
  theme_bw() +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values=color_scales$methods, na.value="grey") +
  theme(legend.position = "top")
```

```{r include=FALSE}
signal_noise_res %>%
  select(dataset, method, true_cell_type, signal_ratio) %>%
  group_by(dataset, method, true_cell_type) %>%
  summarise(median_signal_ratio = median(signal_ratio)) %>%
  spread(true_cell_type, median_signal_ratio) %>%
  write_tsv("../results/tables/spillover_signal_noise.tsv", na="")

```

## Summary figure: migration charts

Download this figure as high quality [pdf](files/spillover_migration_all.pdf)

```{r, fig.width = 10, fig.height=24, echo=FALSE, results='hide', fig.cap="migration charts for all three dataset. The value in the middle of the chart indicates the 'noise ratio', i.e. the sum of all false positive predictions." }
methods = config$deconvolution_methods
layout(matrix(seq(1, length(datasets) * length(methods)), length(methods), length(datasets)))
par(mar=rep(0.5, 4))
circos.par(cell.padding = rep(0, 4))
x = lapply(names(datasets), function(dataset) {
  lapply(methods, function(method) {
    migration = all_results2 %>%
        filter(method == !!method, dataset == !!dataset) %>%
        filter(true_cell_type != "Neutrophil") %>%
        group_by(method, true_cell_type, cell_type) %>%
        summarise(estimate = mean(estimate)) %>%
        ungroup()

    migration_mat = migration %>%
        select(-method) %>%
        spread(cell_type, estimate) %>%
        as.data.frame() %>%
        column_to_rownames("true_cell_type") %>%
        as.matrix()

    noise_ratio = migration %>%
      # aggregate macro/mono into a single category
      mutate(true_cell_type = ifelse(true_cell_type %in% c("Macrophage M1", "Macrophage M2", "Monocyte"), "Macrophage/Monocyte", true_cell_type)) %>%
      mutate(true_cell_type = ifelse(true_cell_type == "T cell regulatory (Tregs)", "T cell CD4+", true_cell_type)) %>%
      group_by(method, cell_type, true_cell_type) %>%
      summarise(estimate = mean(estimate)) %>%
      group_by(method) %>%
      mutate(type = ifelse(cell_type == true_cell_type, "signal", "noise")) %>%
      group_by(method, type) %>%
      summarise(estimate = sum(estimate)) %>%
      spread(type, estimate) %>%
      mutate(noise_ratio = noise/(signal+noise)) %>%
      ungroup()

    chordDiagram(migration_mat, directional = TRUE, transparency = .5,
                 grid.col = color_scales$immune_cells
                 # annotationTrack = c("track", "grid")
                 )

    text(0, 0, method, cex = 2.3)
    text(0, -0.2, as.character(round(filter(noise_ratio, method == !!method) %>% pull(noise_ratio), 2)), cex=2)

    # first method, add title.
    if(method == "mcp_counter") {
      title(dataset)
    }
  })
})

p = recordPlot()
pdf("../results/figures/spillover_migration_all.pdf", width=10, height=24)
p
dev.off()
```
