## Investigate marker genes
Spillover appears to primarily happen between CD8+ and CD4+ T cells and from
DCs to monocytes/macrophages and B cells. We were particularly concerned by the
spillover between DCs and B cells as we could not validate it in the 'immune
cell reference' dataset. In this section, we demonstrate that the single cell
populations of DCs and B cells are both distinct and well-defined and, next,
identify a handful of marker genes that are specific for both plasmacytoid DCs
and B cells and drive the spillover.

```{r, include=FALSE}
mat = cbind(pData(single_cell_schelker$eset), t(exprs(single_cell_schelker$eset))) %>% as_tibble()
mat = mat %>% filter(source != "pbmc")
```


First, we suspected that the effect might be due to misannotations in the single cell dataset.
Here, we show the B cells and DCs in the t-SNE plot of the single cell dataset.
```{r, fig.cap="B cell and DC clusters in the single cell dataset. ", echo=FALSE}
p1 = mat %>%
  mutate(cell_type = ifelse(!cell_type %in% c("Dendritic cell", "B cell"), "other", cell_type)) %>%
  ggplot(aes(x=tsneX1, y=tsneX2, colour=cell_type)) +
           geom_point(size=1) +
           theme(legend.position="right") +
           scale_color_brewer(type="qual", palette=2) +
           guides(colour=guide_legend(ncol = 1, override.aes= list(size=4))) +
           xlab("tSNE-1") +
           ylab("tSNE-2") +
           labs(color="cell type")
```
We observe that both clusters are distinct and non-overlapping.

Next, we look for the expression of typical marker genes of DCs and B cells.
```{r, fig.width=5, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
plots = foreach(gene = c("MS4A1", "CD19", "ITGAM", "ITGAX", "CLEC4C", "IL3RA", "TCL1A"),
        cell_type = c("B cell", "B cell", "myleoid", "myleoid", "plasmacytoid DCs", "plasmacytoid DCs", "plasmacytoid/B")) %do% {
          mat %>% ggplot(aes(x=tsneX1, y=tsneX2)) +
            geom_point(aes_string(colour=gene), size=.4) +
            scale_colour_distiller(palette="RdYlBu", trans="log1p") +
            ggtitle(paste(gene, "-", cell_type))
        }
```


```{r, fig.width=16, fig.height=13}
plot_grid(plotlist = c(list(p1), plots), labels = "auto")

```

**Conclusion**: The B cell and DC cluster separate well, the marker genes do not overlap.
The DC cluster are plasmacytoid DCs, mDCs are somewhere hidden in the Macrophage/Monocyte cluster
and cannot be distinguished.


## Check expression of markers in "detection limit" simulation dataset.
```{r, include=FALSE}
show_cell_types = c("B cell", "Dendritic cell", "Macrophage/Monocyte",
                    "NK cell", "T cell CD4+", "T cell CD8+")
```

```{r, echo=FALSE}
n_immune_cells = rep(c(seq(0, 50, 5)
                     , seq(60, 100, 10)
                     , seq(150, 500, 50)
                     , seq(600, 1000, 100)
                     ), 5)
```


```{r, cache=TRUE, include=FALSE}
alldata = foreach(celltype = names(sensitivity_analysis_dataset),
        .combine=bind_rows) %do% {
          sensitivity_analysis_dataset[[celltype]] %>% as.tibble(rownames="gene") %>% mutate(celltype=celltype)

        }
colname2ncell = data.frame(sample=colnames(sensitivity_analysis_dataset$`B cell`), ncell=n_immune_cells)
alldata2 = alldata %>% gather(sample, TPM, -gene, -celltype)
alldata2 %<>% inner_join(colname2ncell)
```


```{r}
markers = list(
  "B cell" = c("MS4A1", #=CD20
               "CD19", "CD22"
  ),
  "myleoid" = c("ITGAM", "ITGAX"), #=CD11B,CD11C
  "MHC class II" = c("HLA-DRB1", "HLA-DRA"),
  "pDC" = c("CLEC4C", "IL3RA"))
```

We check the abundance of the marker genes in the simulated samples of the
dataset we used for assessing the false positive predictions and the detection
limit.
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Correlation of marker genes with increase of the amount of a certain cell type. Along x-axis: cell type used for simulation. Along y-axis: marker genes. ", fig.height=12, fig.width=12}
markers_df = foreach(cell_type = names(markers), genes=markers, .combine=bind_rows) %do% {
  data.frame(gene=genes) %>% mutate(markers_for=cell_type)
}
allmarkers = unlist(markers)
alldata2 %>% filter(gene %in% allmarkers) %>%
  inner_join(markers_df) %>%
  ggplot(aes(x=ncell, y=TPM)) +
    geom_point(aes(colour=gene)) +
    facet_grid(markers_for+gene~celltype, scales = "free_y") +
    scale_color_brewer(palette="Set3") + stat_cor() + 
    panel_border() + 
    scale_x_continuous(breaks = c(250, 750))
```

```{r, echo=FALSE, fig.width=3, fig.height=3, fig.cap="Even though the number of dendritic cells does correlate with MS4A1, a B cell marker, the absolute expression is really low. This effect can be driven by few misclassified cells but cannot explain the spillover effects we observe. "}

alldata2 %>% filter(celltype == "Dendritic cell", gene == "MS4A1") %>%
  ggplot(aes(x=ncell, y=TPM)) +
  geom_point()+
  stat_smooth(method="lm") +
  stat_cor() +
  ggtitle("MS4A1")
```

### Expression of marker genes in the cell populations
For EPIC, quanTIseq and CIBERSORT, we retreive a list of all genes included in
the signature matrices.

```{r, message=FALSE, warning=FALSE}
# quanTIseq
til10 = read_tsv("../immunedeconv/inst/extdata/quantiseq/TIL10_signature.txt") %>% as.data.frame() %>%
  column_to_rownames("ID")

# EPIC
tref = EPIC::TRef$refProfiles[EPIC::TRef$sigGenes,]

# CIBERSORT
if("cibersort" %in% config$deconvolution_methods) {
  lm22 = read_tsv("../lib/CIBERSORT/LM22.txt") %>% as.data.frame() %>% column_to_rownames("Gene symbol")
}
```


For each marker gene, we show the expression in all cell type populations in
the single cell dataset. We sort the list by the product of the mean expression
in B cells and Dendritic cells to obtain a ranked list of genes
that are enriched in both B cells and DCs. i

```{r, include=FALSE, fig.width=10, fig.height=15, warning=FALSE, message=FALSE, fig.cap="Expression across cell types. Along x-axis: cell types used for simulation. Along y-axis: Signature genes from EPIC/quanTIseq for the cell types. "}
matrices = list(quantiseq=til10, epic=tref)
if(exists("lm22")) matrices$cibersort = lm22
foreach(method = names(matrices)) %do% {
  markers = rownames(matrices[[method]])
  alldata3 = alldata2 %>% filter(gene %in% markers) %>%
    filter(ncell > 600)
  gene_rank = alldata3 %>%
    spread(celltype, TPM) %>%
    mutate(rank = `B cell` * `Dendritic cell`) %>%
    group_by(gene) %>%
    summarise(rank = mean(rank)) %>%
    arrange(rank)

  alldata3 %>%
    ggplot(aes(x=factor(gene, levels=gene_rank$gene), y=TPM)) +
      geom_boxplot() +
      coord_flip() +
      facet_grid(.~celltype, scales = "free", space="free") +
      scale_color_brewer(palette="Set1") +
      theme(strip.text.y = element_text(angle=0)) +
      ggtitle(method) # + scale_y_log10()
  ggsave(paste0("../results/figures/marker_gene_expression_", method, ".pdf"), width = 15, height = ifelse(method=="cibersort", 75, 20), limitsize=FALSE)
}

```

The plots are very large, that's why we don't include them directly.
You can view the plots in full quality as PDF under the following links:

* [EPIC](figures/marker_gene_expression_epic.pdf)
* [quanTIseq](figures/marker_gene_expression_quantiseq.pdf)
* [CIBERSORT](figures/marker_gene_expression_cibersort.pdf)

The figures show the expression of marker genes across cell types. Along x-axis: cell types used for simulation. Along y-axis: signature genes for the cell types derived from the signature matrices of the deconvolution methods.


## Deconvolution results before and after
We checked the most promising candidate genes on
[genecards.org](https://www.genecards.org), and took note of those that
were indeed annotated as being specific for both B cells and plasmacytoid
dendritic cells in the LifeMap Discovery database.

```{r}
B_pDC_genes = c("TCL1A", "TCF4", "CD37", "SPIB", "BCL11A", "IRF8")
```

To assess the effects on deconvolution, we generate
10 simulated dendritic cell samples, run the deconvolution once
with default options and once with the marker genes removed, and
check the 'spillover' to B cells.
```{r, include=FALSE}
set.seed(42)
dc_expression = foreach(i=1:10,
        .combine = bind_cols,
        .final = as.matrix) %do% {
          cell_n = make_cell_fraction_vector("Dendritic cell", k=500, background=NULL)
          cell_frac = cell_n/sum(cell_n)
          make_random_bulk(single_cell_schelker$eset, cell_frac, n_cells=500)
        }
rownames(dc_expression) = rownames(exprs(single_cell_schelker$eset))

# intersection because CIBERSORT (or other methods) might not be available. 
use_methods = intersect(c("epic", "quantiseq", "cibersort"), config$deconvolution_methods)
show_cell_types = c("B cell", "Dendritic cell", "Macrophage/Monocyte",
                    "NK cell", "T cell CD4+", "T cell CD8+")
timer_indications = rep("OV", ncol(dc_expression))

res_before = foreach(method=use_methods,
                     .combine=bind_rows) %do% {
                       deconvolute(dc_expression, method=method, indications=timer_indications) %>%
                         map_result_to_celltypes(show_cell_types, method) %>%
                         as_tibble(rownames = "cell_type") %>%
                         mutate(method=method, dataset="before")
                     }

res_after = foreach(method=use_methods,
                     .combine=bind_rows) %do% {
                       deconvolute(dc_expression, method=method, indications=timer_indications, rmgenes=B_pDC_genes) %>%
                         map_result_to_celltypes(show_cell_types, method) %>%
                         as_tibble(rownames = "cell_type") %>%
                         mutate(method=method, dataset="after")
                     }

res_all = bind_rows(res_before, res_after) %>%
  gather(sample, predicted_fraction, -cell_type, -method, -dataset) %>%
  mutate(dataset = factor(dataset, levels=c("before", "after")))

res_spillover$rm_marker_genes = res_all

```

```{r, fig.width=12, echo=FALSE, message=FALSE, fig.cap="Predictions on 10 simulated DC samples before and after removal of the six genes. ", warning=FALSE}
res_all %>%
  ggplot(aes(x=dataset, y=predicted_fraction, colour=dataset)) +
    geom_quasirandom() +
    stat_summary(fun.y=mean, geom="crossbar", fun.ymin=mean, fun.ymax=mean, width=.5, color="black") +
   facet_grid(method~cell_type, drop = TRUE, scales = "free") +
  stat_compare_means(paired = TRUE, method = "t.test")


```
