```{r setup, include=FALSE}
# tidyverse
library(dplyr)
library(dbplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(magrittr)
library(readxl)
library(stringr)
# other packages
library(cowplot)
library(pheatmap)
library(ggpubr)
library(DBI)
library(testit)
library(knitr)

devtools::load_all("../immunedeconv")

color_file = "../tables/color_scales.xlsx"
color_scales = sapply(excel_sheets(color_file), function(sheet) {
  tbl = read_excel(color_file, sheet=sheet)
  colors = tbl$color
  names(colors) = tbl$value
  colors
})

tcga_db = dbConnect(RSQLite::SQLite(), "/db/tcga.sqlite", flags = RSQLite::SQLITE_RO)
```

# Correlation of methods across TCGA samples. 

Next, we were interested how the predictions for different cell types correlate. 
To this end, we ran all methods on `r tbl(tcga_db, "sample") %>% collect() %>% nrow()` TCGA samples. 

If predictions for the same cell type of different methods correlate strongly, we 
can assume the signatures to be robust. If they do not cluster, 
the signatures are different and we can conclude, that potentially the cell-type is not well-defined. 

The data has been preprocessed elsewhere, and we obtain the estimates from our database. 

```{r, cache=TRUE, echo=FALSE}
# Obtain the data from database
tcga_infiltration = tbl(tcga_db, "infiltration") %>%
  collect() %>% 
  filter(method != "cibersort") %>% # cibersort does not make senes here, as it only compares relative fractions. 
  mutate(method_cell_type = paste0(cell_type, " [", method, "]")) %>%
  select(sample, method_cell_type, estimate) %>%
  spread(method_cell_type, estimate) %>% 
  na.omit() # remove all samples that do not have estimates by all methods (LAML does not have TIMER estimates. )

tcga_infil_mat = tcga_infiltration %>%
  select(-sample) %>% 
  as.matrix()

tcga_cor = cor(tcga_infil_mat)
```

Analysis is based on `r tcga_infiltration %>% select(sample) %>% distinct() %>% nrow()` samples that have estimates by all methods. 

## Heatmap clustered by correlation
```{r, fig.width=12, fig.height=12, message=FALSE}
annot_df = str_match(colnames(tcga_cor), "(.*) \\[(.*)\\]") %>% as.data.frame() 
colnames(annot_df) = c("names", "cell type", "method") 
annot_df %<>% column_to_rownames("names")

use_cell_types = names(color_scales$cell_types)

immunedeconv::map_cell_types(use_cell_types, )

tcga_clustered = pheatmap(tcga_cor, 
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         annotation_row = annot_df)

# ggsave("../results/figures/tcga_correlation_clustered.pdf", plot=tcga_clustered$gtable)
# ggsave("../results/figures/tcga_correlation_clustered.png", plot=tcga_clustered$gtable, dpi=1200)
```

## Heatmap unclustered, sorted by cell type
```{r, fig.width=12, fig.height=12}
pheatmap(tcga_cor,
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```